Binning Strategies and Related Loss for Large Data

### 2015-05-08 Paper.Rnw
added reference

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. Binned scatterplots are inherently biased displays of spatial information as they shift visual emphasis from the true location of individual points to the geometric centers of tiles. However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
\ktm{When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator. We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator. It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss. This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies to bin shades. %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.  
}

While bin widths can be set to any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual attributes in the binned scatterplot. If the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins throughout the binned scatterplot.  These white-space gaps are undesirable because they create visual discontinuity that interferes with interpretation of bivariate density. There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.

>>>>

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. Binned scatterplots are inherently biased displays of spatial information as they shift visual emphasis from the true location of individual points to the geometric centers of tiles. However, as bin sizes become increasingly fine,  the bias decreases as the plot shows more precise spatial locations.
When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate is spatially exact; thus making the binned scatterplot a consistent visual estimator. We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator. It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss. This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies. %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.

While bin widths can be chosen as any positive real value, the selection should be restricted to an integer multiple of the resolution of the data to avoid undesirable visual artefacts in the binned scatterplot. For example, if the bin dimensions are smaller than the resolution of the data, there will be empty rows or columns of bins in the binned scatterplot.  These gaps of white space are undesirable because they create visual discontinuity that interferes with the interpretation of bivariate density. 
\hh{The grid of white spaces also creates an optical illusion, the Hermann-grid illusion~\citep{hermann:1870, spillmann:1994}, which makes dark spots appear in the crossings of the lines and additionally interferes with an evaluation.}

There is also a disruptive visual consequence if bin dimensions are not integer multiples of the bivariate data resolution; where bins have systematically different numbers of possible data values that can cause \textit{artificial striping} -- an oscillating density pattern imposed by the binning that does not exist in the raw data -- to appear in the binned scatterplot.
